{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-19 thesis/tm.tex
tm map basicer mutex 2,4
finish the last tm test
daves final revisions to tm

Transactional memory \revisionminor{(TM)~\cite{transactional-memory}} is a concurrency primitive
by which programmers may attempt an arbitrary sequence of shared memory accesses,
which will either be committed (made visible to other processors/threads) atomically,
such that no intermediate state modification is ever visible,
or, in the case of a conflict which would prevent such,
discarded with an error code returned to allow the programmer to \revisionminor{invoke} a synchronized backup path.
%
Transactional memory specifications typically have three API functions, abstractly speaking:
\begin{itemize}
	\item  {\sf \revisionminor{begin}} begins a transaction,
		staging any subsequent shared memory accesses in some temporary thread-/processor-local storage,
		and checking for conflicts with the accesses of any other threads or CPUs.
		\revisionminor{When the transaction is started, {\sf begin} returns a success code.}
		If the transaction is unsuccessful, as described below,
		{\sf begin} instead returns an error code
		indicating the programmer should fall back to some other, possibly slower, synchronization method.
	\item {\sf \revisionminor{end}} ends a transaction,
		attempting to commit all staged accesses to the shared memory atomically with respect to
		reads or writes from other concurrently-executing code.
		If any of those accesses conflict (i.e., read/write or write/write)
		with any other access to the same memory since the transaction started,
		they are instead discarded and execution state reverts to the {\sf begin} with an error code as described above.
	\item {\sf \revisionminor{abort}} explicitly aborts a transaction,
		regardless of any memory conflicts,
		discarding changes and reverting execution as described above.
		Some implementations allow an arbitrary abort code to be specified
		which will appear \revisionminor{as} {\sf begin}'s error code.
\end{itemize}

@@

\revision{For the most part,
existing stateless model checkers
model interrupt-driven thread switching as the only source of nondeterminism in a program's execution.
In other words, assuming a fixed test input,%
\footnote{Symbolic execution \cite{symbolic}, a neighbouring research area,
addresses input nondeterminism; see \cref{related:symbolic}.}
the only way multiple behaviours can arise is by scheduling different threads at each preemption point.
Only recently have checkers emerged that extend this model by
incorporating other nondeterminism sources:
store buffer reordering arising from relaxed memory models
\cite{cdschecker,tsopso,rcmc}
and event-driven application programming models
\cite{r4,eventracer},
both of which are especially relevant in mobile programming.
Under TM, {\sf begin}'s return value is also nondeterministic,
and moreover may depend on execution events from the future in a parallel universe
(i.e., subsequent memory conflicts detected and reverted),
and no model checker yet exists which can model this.
In this chapter I will develop a concurrency model to incorporate abort nondeterminism
(\cref{sec:tm-design}),
including two equivalence proofs to help cope with the consequent state space explosion
(\cref{sec:tm-design-failure} and \cref{sec:tm-design-locks}),
extend the proofs to account for weak atomicity (\cref{sec:tm-weak-atomicity}),
extend Landslide to support testing TM programs under a variety of execution semantics
(\cref{sec:tm-implementation}),
and demonstrate its practicality
%\cref{sec:tm-eval}
both by finding new bugs
(\cref{sec:tm-eval-bugs})
and by verifying correctness
(\cref{sec:tm-verif})
in a variety of microbenchmarks and real-world HTM programs.}

@@

\begin{figure}[t]
	\begin{center}
		\begin{tabular}{ll}
			\hilight{lavender}{\bf Thread 1} &
			\hilight{seafoam}{\bf Thread 2} \\
			\hline
			\texttt{status = \call{\_xbegin}()} & \\
			\texttt{temp <- x;} & \\
			\texttt{temp <- temp + 1;} & \\
			& \texttt{status = \call{\_xbegin}()} \\
			& \texttt{x++;} \\
			& \texttt{\call{\_xend}();} \\
			\texttt{x <- temp;} & \\
			\texttt{\call{\_xend}();} & \\
		\end{tabular}
	\end{center}
	\caption[Impossible hypothetical interleaving within transactions of \Cref{fig:htm-example-assemblified}.]
		{Impossible hypothetical interleaving within transactions of \Cref{fig:htm-example-assemblified}.
		The effects of $\tiat{2a-2b}$ are never globally visible, as its transasction must abort,
		and moreover $\tiat{2c-3}$ will not execute at all.}
	\label{fig:htm-example-interleaving}
\end{figure}
}

@@

\Cref{thm:atom} must be weakened correspondingly.
\vspace{1em}

@@

@@

			& 3,3 & \ETAdag{qqq} & \ETAdag{qqq} \\

@@

			& 4,2 & \ETAdag{qqq} & \ETAdag{qqq} \\

@@

			& 2,4 & \ETAdag{qqq} & \ETAdag{qqq} \\

@@

\begin{itemize}
	\item Three equivalence proofs,
		demonstrating that model checkers can soundly emulate strongly-atomic HTM execution semantics by
		treating transactions as globally-uninterruptible atomic sections
		and injecting failure without simulating rollback
		(\cref{sec:tm-design-failure})
		and applying conventional data race analysis with all transactions treated as a global lock
		(\cref{sec:tm-design-locks}),
		and can emulate weakly-atomic STM systems by relaxing the above atomicity to apply to other transactions only
		(\cref{sec:tm-weak-atomicity}).
	\item An implementation of both forms of transactional memory in Landslide,
		with further options to configure the variety of abort codes injected
		(\cref{sec:tm-implementation}).
	\item A suite of transactional benchmark programs,
		designed to produce interesting state spaces of varying size, bugginess, and reducibility
		(\cref{sec:tm-eval-exp-setup}).
		This test suite is more appropriate for evaluating model checkers
		% get rekt STAMP :triumph:
		than prior work, which to date has focused on testing the raw performance of TM backend implementations,
		with little focus on the interactions of abort paths.
	\item Two previously-unknown bugs found in transactional programs from prior work
		(\cref{sec:tm-eval-bugs}).
	\item Retry set reduction,
		a new state space reduction algorithm
		(\cref{sec:tm-retrysets})
		which the evaluation shows identifies and skips equivalent interleavings
		compared to standard DPOR, even when transactional and abort paths are fully mutually conflicting
		(\cref{sec:tm-eval-retrysets}).
	\item Demonstration of two further heuristic reduction strategies on HTM programs,
		abort code reduction (\cref{sec:tm-eval-stm})
		and abstraction reduction (\cref{sec:tm-abstraction})
		which rely on human intuition to ensure their soundness
		to tackle even larger state spaces still.
	\item Verifications of \Cref{fig:htm-fixed}'s atomicity protocol %(with retry set reduction)
		and the {\tt lock}({\tt mutex}) transactional lock %(with abort code reduction)
		up to 2 threads, 4 iterations and 4 threads, 1 iteration for each
		(\cref{sec:tm-verif}).
		%$(K,N)$=$(2,4)$ and $(4,1)$ for each.
		% verified htm atomicity protocol (htm2) and mutex (lock mutex) up to (2,4), (4,1)
\end{itemize}

>>>>

Transactional memory \revisionminor{(TM)~\cite{transactional-memory}} is a concurrency primitive
by which programmers may attempt an arbitrary sequence of shared memory accesses,
which will either be entirely committed (i.e., made visible to other processors/threads) atomically,
such that no intermediate state modification is ever visible,
or, in the case of a conflict which would prevent such,
entirely discarded with an error code returned to allow the programmer to \revisionminor{invoke} a synchronized backup path.
%
Transactional memory specifications typically have three API functions, abstractly speaking:
\begin{itemize}
	\item  {\sf \revisionminor{begin}} begins a transaction,
		staging any subsequent shared memory accesses in some temporary thread-/processor-local storage,
		and checking for conflicts with the accesses of any other threads or CPUs.
		\revisionminor{When the transaction is started, {\sf begin} returns a success code.}
		If the transaction is unsuccessful, as described below,
		{\sf begin} instead returns an error code
		indicating the programmer should fall back to some other, possibly slower, synchronization method.
	\item {\sf \revisionminor{end}} ends a transaction,
		attempting to commit all staged accesses to the shared memory atomically with respect to
		reads or writes from other concurrently-executing code.
		If any of those accesses conflict (i.e., read/write or write/write)
		with any other access to the same memory since the transaction started,
		they are instead discarded and execution state reverts to the {\sf begin} point
		with an error code as described above.
	\item {\sf \revisionminor{abort}} explicitly aborts a transaction,
		regardless of any memory conflicts,
		discarding changes and reverting execution as described above.
		Some implementations allow an arbitrary abort code to be specified
		which will appear \revisionminor{as} {\sf begin}'s error code.
\end{itemize}

@@

\revision{For the most part,
existing stateless model checkers
model interrupt-driven thread switching as the only source of nondeterminism in a program's execution.
In other words, assuming a fixed test input,%
\footnote{Symbolic execution \cite{symbolic}, a neighbouring research area,
addresses input nondeterminism; see \cref{related:symbolic}.}
the only way multiple behaviours can arise is by scheduling different threads at each preemption point.
Only recently have checkers emerged that extend this model by
incorporating other nondeterminism sources:
store-buffer reordering arising from relaxed memory models
\cite{cdschecker,tsopso,rcmc}
and event-driven application programming models
\cite{r4,eventracer},
both of which are especially relevant in mobile programming.
Under TM, {\sf begin}'s return value is also nondeterministic,
and moreover may depend on execution events from the future in a parallel universe
(i.e., subsequent memory conflicts detected and reverted),
and no model checker yet exists which can model this.
In this chapter I will develop a concurrency model to incorporate abort nondeterminism
(\cref{sec:tm-design}),
including two equivalence proofs to help cope with the consequent state space explosion
(\cref{sec:tm-design-failure} and \cref{sec:tm-design-locks}),
extend the proofs to account for weak atomicity (\cref{sec:tm-weak-atomicity}),
extend Landslide to support testing TM programs under a variety of execution semantics
(\cref{sec:tm-implementation}),
and demonstrate its practicality
%\cref{sec:tm-eval}
both by finding new bugs
(\cref{sec:tm-eval-bugs})
and by verifying correctness
(\cref{sec:tm-verif})
in a variety of microbenchmarks and real-world HTM programs.}

@@

\begin{figure}[t]
	\begin{center}
		\begin{tabular}{ll}
			\hilight{lavender}{\bf Thread 1} &
			\hilight{seafoam}{\bf Thread 2} \\
			\hline
			\texttt{status = \call{\_xbegin}()} & \\
			\texttt{temp <- x;} & \\
			\texttt{temp <- temp + 1;} & \\
			& \texttt{status = \call{\_xbegin}()} \\
			& \texttt{x++;} \\
			& \texttt{\call{\_xend}();} \\
			\texttt{x <- temp;} & \\
			\texttt{\call{\_xend}();} & \\
		\end{tabular}
	\end{center}
	\caption[Impossible hypothetical interleaving within transactions of \Cref{fig:htm-example-assemblified}.]
		{Impossible hypothetical interleaving within transactions of \Cref{fig:htm-example-assemblified}.
		The effects of $\tiat{2a-2b}$ are never globally visible, as its transaction must abort,
		and moreover $\tiat{2c-3}$ will not execute at all.}
	\label{fig:htm-example-interleaving}
\end{figure}
}

@@

\Cref{thm:atom} must be weakened correspondingly, yielding \Cref{thm:atom-weak}.
\vspace{1em}

@@

			& 2,5 & \cpu{3417.86} & \ints{76787} \\

@@

			& 3,3 & \ETAdag{11d 22h} & \ETAdag{23136172} \\
			% [JOB 0] progress: 764124/23136172 brs (3.302724%), ETA 11d 11h 45m 28s (elapsed 9h 59m 59s)

@@

			& 4,2 & \ETAdag{41d 0h} & \ETAdag{66520074} \\
			% [JOB 0] progress: 702872/66520074 brs (1.056631%), ETA 40d 13h 44m 41s (elapsed 9h 59m 59s)

@@

			& 2,4 & \ETAdag{2d 17h} & \ETAdag{1130184} \\
			% [JOB 3] progress: 572848/1130184 brs (50.686221%), ETA 2d 6h 52m 20s (elapsed 9h 59m 55s)

@@

\begin{itemize}
	\item An approach for model checkers to emulate strongly-atomic HTM
		by treating transactions as globally-uninterruptible atomic sections
		and injecting failure without simulating rollback
		(\cref{sec:tm-design-failure})
		and applying conventional data race analysis with all transactions treated as a global lock
		(\cref{sec:tm-design-locks}),
		plus an extension to also emulate weakly-atomic STM systems
		by relaxing the above atomicity to apply to other transactions only
		(\cref{sec:tm-weak-atomicity}),
		each section with a corresponding soundness proof of equivalence to existing systems' semantics.
	\item An implementation of both forms of transactional memory in Landslide,
		with further options to configure the variety of abort codes injected
		(\cref{sec:tm-implementation}).
	\item A suite of transactional benchmark programs,
		designed to produce interesting state spaces of varying size, bugginess, and reducibility
		(\cref{sec:tm-eval-exp-setup}).
		This test suite is more appropriate for evaluating model checkers
		% get rekt STAMP :triumph:
		than prior work, which to date has focused on testing the raw performance of TM backend implementations,
		with little focus on the interactions of abort paths.
	\item Two previously-unknown bugs found in transactional programs from prior work
		(\cref{sec:tm-eval-bugs}).
	\item Retry set reduction,
		a new state space reduction algorithm
		(\cref{sec:tm-retrysets})
		which the evaluation shows identifies and skips equivalent interleavings
		compared to standard DPOR, even when transactional and abort paths are fully mutually conflicting
		(\cref{sec:tm-eval-retrysets}).
	\item Demonstration of two further heuristic reduction strategies on HTM programs,
		abort code reduction (\cref{sec:tm-eval-stm})
		and abstraction reduction (\cref{sec:tm-abstraction})
		which rely on human intuition to ensure their soundness
		to tackle even larger state spaces still.
	\item Verifications of \Cref{fig:htm-fixed}'s atomicity protocol %(with retry set reduction)
		and the {\tt lock}({\tt mutex}) transactional lock %(with abort code reduction)
		up to 2 threads, 4 iterations and 4 threads, 1 iteration for each
		(\cref{sec:tm-verif}).
		%$(K,N)$=$(2,4)$ and $(4,1)$ for each.
		% verified htm atomicity protocol (htm2) and mutex (lock mutex) up to (2,4), (4,1)
\end{itemize}
